---
title: "Diversity report"
author: "CSB5 Shotgunmetagenomics-nf"

output:
  html_document:
    df_print: paged
    code_folding: hide

params:
  metadata: NA
  profile_list: NA
  presence_absence_thre: 0.1
  index_column: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Dependencies

Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(ggplot2)
library(vegan)
library(dplyr)
library(purrr)
library(readr)
library(stringr)
library(plotly)
```

Utility constants and functions defined
```{r functions}
## A set of colors
color.list <- c("#E64B35FF", "#4DBBD5FF", "#00A087FF", "#3C5488FF", "#F39B7FFF", "#8491B4FF",
                "#91D1C2FF", "#B09C85FF", "#FAFD7CFF", "#82491EFF", "#B7E4F9FF", "#FB6467FF",
                "#526E2DFF", "#E762D7FF", "#FAE48BFF", "#A6EEE6FF", "#95CC5EFF")
## ggplot2 preset theme
figtheme <- theme_classic() +
  theme(text = element_text(size=23,face='bold'),
        axis.title.y=element_text(margin=margin(0,15,0,0)),axis.title.x=element_text(margin=margin(15,0,0,0)),
        plot.margin = unit(c(1,1,1,1), "cm"),
        plot.title = element_text(margin=margin(0,0,15,0), hjust=0.5))
theme_set(figtheme)
## Auxilary function to merge two profiles
merge2 <- function(x, y){
  aux <- function(x) read_tsv(x, col_names=c("tax", str_remove(basename(x), '[\\._\\-].*')))
  if(is.character(x)){
    dat.x <- aux(x)
  }else{
    dat.x <- x
  }
  dat.y <- aux(y)
  full_join(dat.x, dat.y)  
}
```

# Load data

```{r load, message=FALSE, cols.print=6, rows.print=10}
profile.list <- read_lines(params$profile_list)
tax.profile <- reduce(profile.list, merge2)
tax.profile <- tibble::column_to_rownames(tax.profile, 'tax')

tax.profile[is.na(tax.profile)|tax.profile<params$presence_absence_thre] <- 0 ## impute NAs and set low abundance taxon to 0
tax.profile <- tax.profile[rowSums(tax.profile) != 0, ] ## remove empty taxa
tax.profile

if(params$metadata!='NA'){
  metadata <- read_tsv(params$metadata)
  metadata
}
```
# $\alpha$-diversity overview 

## Shannon diversity {.tabset}
Shannon diversity for a sample with $S$ taxa:

$$H=-\sum_{i=1}^S p_iln(p_i)$$

### Overview

```{r message=FALSE}
div <- diversity(tax.profile, index='shannon', MARGIN=2)
p <- ggplot(data.frame(div=div), aes(x=div))+
  geom_histogram() + 
  labs(x='Shannon index', y='Count')
ggplotly(p)
```

### Sample specific

```{r message=FALSE, fig.height=6, fig.width=20}
p <- 
  ggplot(data.frame(div=div, Sample=colnames(tax.profile)), 
         aes(y=div, x=reorder(Sample, div)))+
  geom_bar(stat='identity') + 
  labs(y='Shannon index', x='Sample') + 
  theme(axis.text.x = element_blank())
ggplotly(p)
```

## Simpson diversity {.tabset}

Simpson diversity for a sample with $S$ taxa:

$$D=\frac{1}{\sum_{i=1}^Sp_i^2}$$

### Overview
```{r message=FALSE}
div <- diversity(tax.profile, index='simpson', MARGIN=2)
p <- ggplot(data.frame(div=div), aes(x=div))+
  geom_histogram() + 
  labs(x='Simpson index', y='Count')
ggplotly(p)
```

### Sample specific
```{r message=FALSE, fig.height=6, fig.width=20}
p <- 
  ggplot(data.frame(div=div, Sample=colnames(tax.profile)), 
         aes(y=div, x=reorder(Sample, div)))+
  geom_bar(stat='identity') + 
  labs(y='Simpson index', x='Sample') + 
  theme(axis.text.x = element_blank())
ggplotly(p)
```

# $\beta$-diversity overview (Bray-Curtis dissimilarity) 

Bray-Curtis Dissimilarity between sample $i$ and $i'$ with $P$ taxa:
$$Bray_{ii'}=\frac{\sum_{j=1}^P|x_{ij}-x_{i'j}|}{\sum_{j=1}^Px_{ij}+x_{i'j}}$$

```{r beta_pcoa}
distance <- vegdist(t(tax.profile))
cmds <- cmdscale(distance, eig = TRUE)
perc <- (cmds$eig/sum(cmds$eig))[1:2]*100
plot.dat <- data.frame(cmds$points)
if(exists("metadata")){
  plot.dat <- merge(plot.dat, metadata, by.x=0, by.y=params$index_column, all.x=TRUE)
}
```

## Overview
```{r beta_pcoa_plot}
## sample code, change "Variable" to the column name in your metadata
p <- ggplot(plot.dat, aes(x=X1, y=X2, col="Variable")) + 
    geom_density_2d(color='grey') + 
    geom_point(size=3) + 
    scale_color_manual(values=color.list) + 
    labs(x=sprintf('PCoA 1 [%.1f%%]', perc[1]), y=sprintf('PCoA 2 [%.1f%%]', perc[2])) 
ggplotly(p)
```

## PCoA plots with associated metadata {.tabset}
```{r echo=FALSE, fig.height=6, fig.width=10, results='asis'}
if(exists("metadata")){
  tmp <- apply(metadata,2,function(x) length(unique(x))) 
  headers <- names(tmp)[tmp<length(color.list) & !sapply(metadata, is.numeric)] ## only support those variables with less category than the number of colors; ignore numeric variables

  for (h in headers){
    cat("###", h, '<br>', '\n')
    cat('PCoA stratefied by: ', h, '<br>')
    p <- ggplot(plot.dat, aes_string(x="X1", y="X2", col=h)) + 
      geom_density_2d(color='grey') + 
      geom_point(size=3) + 
      scale_color_manual(values=color.list, na.value='grey') + 
      labs(x=sprintf('PCoA 1 [%.1f%%]', perc[1]), y=sprintf('PCoA 2 [%.1f%%]', perc[2])) 
    print(p)
    cat('\n', '<br>', '\n\n')
  }
}else{
  print("No metadata given.")
}
```

## Analysis session information

```{r echo=FALSE}
sessionInfo()
```
